Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-07-23 MS/ms.Rmd
Fixed up MS with sd comments

---
title: "Meta-analytic approaches and effect sizes to account for 'nuisance heterogenity' in comparative physiology"
bibliography: ../bib/refs.bib
csl: ../bib/the-journal-of-experimental-biology.csl
output: 
  bookdown::word_document2:
    toc: no
    toc_depth: 6
    number_sections: false
    reference_docx: template.docx
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    toc: no
    toc_depth: 6
    toc_float: yes
    bibliography: refs.bib
editor_options: 
  chunk_output_type: console
---

@@

Meta-analysis is always concerned with effect heterogeneity; in other words, the factors that drive differences in the direction and magnitude of effects within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019]. The concept of heterogeneity is vitally important because it tells us how general our findings are likely to be and how much of the variance we see is the result of real biology or methodological differences (not just sampling variance) [@Nakagawa2017; @Borenstein2019].

There are many phenomena driving effect heterogeneity. First, there is inherent uncertainty in estimating the ‘true’ population effect (e.g., a correlation coefficeiont) from smaller samples [@Borenstein2019]. A formal meta-analysis weights sample estimates by their precision, and removes this source of heterogeneity [@Gurevitch2018; @Koricheva2013]. Removing sampling variance is possible because we know, mathematically, how to calculate it for many effect measures. While unweighted analyses are common, weighting studies makes overall estimates more precise, and less susceptible to publication bias (see below), even if the overall mean itself is unbiased [@Morrissey2016a]. Second, there are sociological factors impacting heterogeneity. These include the ease with which novel and significant results are published relative to non-significant results, manifesting as publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, real biological processes can drive effect heterogeneity. This is particularly true in comparative physiology where we synthesise data from different species with varying life-histories, habitats, mating systems and reproductive modes. As comparative physiologists, this is the stuff that gets us up every morning! We can model these biological factors to test predictions from hypotheses that we believe are key players impacting mean effect size and direction. This improves our understanding of the biological world around us and helps inform on future research directions.

Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it is important we know and understand this to determine how experiments can be designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' [following from @Cooperetal2019] -- factors we know vary, but are rather unsurprising to us (Figure \@ref(fig:fig1)). These might include differences in temperature, pH, dosages, water potential etc. While these factors are likely to have an impact on the magnitude of an effect, they might distract from the core interest of a meta-analysis. In some cases, however, the latter factors might actually not be 'nuisance' variables but the main interest of a synthesis. Regardless, in many cases they can result in a more complicated interpretation of an effect measure than what is desired.

@@

Besides sampling variance, all the sources of heterogeneity discussed above contribute (to varying extents) to the 'apples and oranges' problem often touted as compromising the reliability of meta-analyses. Combining standardised effects, while ignoring these sources of heterogeneity can indeed weaken the reliability of a meta-analysis. However, many argue that heterogeneity is just what we are interested in synthesizing [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. In the words of Gene Glass, the founder of meta-analysis: "Of course it [meta-analysis] mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. Whether comparing apples and oranges is a good idea simply boils down to the specific question of interest, the population one wishes to make conclusions about and how heterogeneous effects actually are. If we are interested in generalising only to apples, then mixing fruit will not be ideal. However, if we want to understand why different 'fruit' vary in their effects, then we need to consider apples and oranges together [@Borenstein2019].

We tend to agree with @Glass2015, and his sentiment is particularly relevant to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms, including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are different. Synthesizing diverse study outcomes allows us to also understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction), and approaches (e.g., short-term vs long-term studies) that can explain the diversity of effect outcomes observed. Doing so provides a rich set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included in a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant and may even be misleading. It may, therefore be more useful to understand overall effects in each of these groups separately, and assess the extent to which they differ.

@@

Before overviewing strategies to deal with nuisance heterogeneity, it is worth first understanding how comparative physiologists are currently conducting meta-analyses and other quantitative syntheses. As such, we conducted a literature survey to better understand the different types of effect sizes being used in comparative physiology, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled. After searching for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species, we we asked: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies (i.e., 'nuisance heterogeneity')? 3) If so, was such variation accounted for in the effect size or during analysis? and finally, 4) how many analyses used 'weighted' meta-analytic models?

@@

We included comparative studies / meta-analyses that were focused on physiological questions. Narrative reviews and comments were not considered because they did not collate quantitative data from the literature. Studies needed to have collected physiological trait data, such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (e.g., heart rates, sprint speed, swim speed, growth rates), body composition (e.g., fatty acid profiles) and immune traits (e.g., cell function, antibodies). Studies were also included if they measured the relationship between physiological traits and body size (e.g., metabolic scaling), reproductive allocation, or survival across taxa. Studies on behavior were only included if the synthesis was explicitly interested in the proximate physiological drivers of behavioral variation. We also included meta-analyses that quantified plasticity in physiological traits in response to environmental variation. While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature because this is the major data source in quantitative syntheses. We excluded quantitative syntheses interested in phenology, bio-mechanics, sexual size dimorphism (i.e., except if the study was interested in sexual differences in physiological traits), those focused on validating physiological methodology, or characterising population genetic structure, gene transcription, and quantitative genetics. There were also meta-analyses that focused on colour evolution which varied in their goals but were most often not focused on physiological questions. As such, these studies were also excluded. We restricted our focus to animal studies because most plant synthesis papers were focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output, and greenhouse gas emissions).

@@

As can be seen in Table \@ref(tab:tablea), none of the effect measures control for 'nuisance heterogeneity'. For example, the magnitude of *lnRR* or *SMD* might depend on differences in the temperatures, dosages, pH, or water potential applied to each treatment. The same applies for modelling the means. For example, standard metabolic rate (SMR) is temperature-dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest itself). While we could model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling (Figure \@ref(fig:fig0)). In Box 1 we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in the box focusing mainly on generality. Below, we show how we can formalise a traditional effect size, such as $lnRR$, to capture this ‘nuisance variation’ using a common, well understood effect measure of interest to comparative physiologists ($Q_{10}$).

```{r, fig0, fig.cap="**Effect sizes from old to new**. We still need a legend..insert here?"}
image <- image_read(here::here("figs", "Fig0_v1_ML.png"))
image
```

@@

Assume that we are interested in understanding the overall effect of, say, corticosterone (CORT) on reproduction [e.g., see @Podmola2018]. We might extract data from studies manipulating corticosterone in bird eggs. Of course, these studies are diverse, spanning many species and applying different dosages of corticosterone. If the studies are experimental in nature, then a standardised effect size, such as $lnRR$ or *SMD*, might be applicable. We then might fit an overall multilevel meta-analytic model as follows: 
$$
\begin{split}
Y_{i} = \beta_{0} + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i} \\
a_{k[i]}\sim N(0, \sigma^{2}_{phylogeny}\textbf{A}), \\
sp_{k[i]}\sim N(0, \sigma^{2}_{species}\textbf{I}), \\
s_{j[i]}\sim N(0, \sigma^{2}_{study}\textbf{I}), \\
e_{i}\sim N(0, \sigma^{2}_{residual}\textbf{I}), \\
m_{i} \sim N(0, v_{i}\textbf{I}) \\
\end{split}
$$

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect size, $v_{i}$ (which is known as the sampling variance for the effect). This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Importantly, if we have effect sizes that are derived from the *same* sample of organisms (e.g., because traits are measured on the same sets of individuals), then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying $\textbf{I}$ to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation, such as differences in the methods applied, species differences, or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our question: across all the species and studies manipulating egg corticosterone, how much of an overall effect does corticosterone have on reproduction relative to control conditions?

Of course, the overall effect does not provide us with a way to interpret its magnitude in the context of how much CORT has been applied in the treatments used in the sample of studies. The most we can say about it is that the effect is small, moderate or large given the sample of effect sizes extracted from this set of studies. In the next section, we show how to 'correct' the overall effect size so that its interpretation is more meaningful.

@@

where $\sum \beta_{j}x_{j}$ is simply the sum of all effects for all moderators (fixed effects), $x_{j}$. All other notation is the same as described above. Importantly, the variable $x_{j}$ could be any moderator collected from studies, including dosage, temperature, salinity, or pH. Some moderators will be at the level of the study (e.g., was the study experimental or observational in nature) but others could be taken at the effect size level (i.e., was the effect derived using a temperature of 23°C or 35°C). By including these moderators in a meta-regression model, the interpretation of the overall meta-analytic mean, $\beta_{0}$, changes. For example, assume we collected the change in CORT dosage relative to a control group (i.e., in µL / g) for our meta-analysis on egg CORT effects on bird reproduction. If the untransformed dosage, $x_{j}$, is included in the model then the overall meta-analytic mean estimate, $\beta_{0}$, would be the mean effect of CORT on reproduction when the dosage difference between the treatment and control group is 0 µL /g. Of course, this interpretation does not make much sense.

We can, however, apply some simple transformations to make the overall mean estimate more intuitive while also improving the interpretation of non-linear parameters and interactions estimated from the model [@Schielzeth2010; @GelmanHill2007]. One simple transformation would be to centre the variable $x_{j}$ by subtracting each value from the mean to create a new input variable (i.e., $c_{i}=x_{i}-\bar{x}$, where the subscript *i* denotes each individual effect size dosage). The new variable, $c_{j}$, is now centered on the mean and can replace $x_{j}$ in the meta-regression model. When the model is refit with $c_{j}$, the estimated $\beta_{0}$ can now be more intuitively interpreted as the overall mean effect of CORT on reproduction at an average dosage difference between treatment and control groups. Importantly, by centering, we retain the original units (µL / g) of the variable, which can be very useful to ease the interpretation of the effect magnitude (e.g., when the dosage difference is 10 µL /g). Given it is very common to estimate and compare overall meta-analytic means (i.e., $B_{0}$) in meta-analyses, this centering can be particularly useful in making these means more interpretable. Mean centering can even be extended to include other 'nuisance' variables that might creep into the dataset; the same centering can be done with a second variable, and the overall mean adjusted and interpreted in the context of this new variable along with dosage. Of course, if one is not interested in the magnitude of the effect, we can simply model mean reproductive output as a function of dosage (often what are referred to as 'arm-based' meta-analytic models). Here, one can model the mean for each group (i.e., control and treatment) and account for the known sampling variance associated with each group mean. Dosage, or other continuous variables, like temperature, can then be modeled (accounting for sampling variance) to understand how mean reproductive output varies. It is important to recognise that all means need to be comparable and that these models are more complex to fit. While they can be equivalent to 'contrast-based' models, they require complex interactions to be estimated [@Nakagawa2015].

Centering moderator variables, like temperature and dosage, not only provides a way to make the overall mean effect more interpretable in the face of treatment heterogeneity, but it also allows more complex relationships to be fit without compromising interpretations, such as when one includes non-linear parameters (e.g., polynomials) and interactions with other variables. In other words, linear / main effects can still be interpreted normally in the presence of non-linear / interactions fit in the model [@Schielzeth2010; @GelmanHill2007]. We provide examples in the online Supplementary Material (with R code) to show readers how the models we describe above can be fit, and how meta-analytic means can be adjusted for treatment heterogeneity.

#### *Heterogeneity Analysis: How Much Effect Size Variation Actually Exists?*

@@

We would like to thank the Editorial board (Craig Franklin, Michaela Handel and Charlotte Rutledge) for inviting us to contribute to this special issue along with feedback on previous drafts. We would also like to thank Megan Head for comments on a previous draft. DWAN, SN, ML and REO were supported by ARC Discovery Projects (DP210101152 to DWAN and DP200100367 to SN, ML and DWAN). PP and SB were supported by a UNSW Scientia PhD Scholarship.

>>>>

---
title: "Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology"
bibliography: ../bib/refs.bib
csl: ../bib/the-journal-of-experimental-biology.csl
output: 
  bookdown::word_document2:
    toc: no
    toc_depth: 6
    number_sections: false
    reference_docx: template.docx
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    toc: no
    toc_depth: 6
    toc_float: yes
    bibliography: refs.bib
editor_options: 
  chunk_output_type: console
---

@@

Meta-analysis is always concerned with effect heterogeneity; in other words, the factors that drive differences in the direction and magnitude of effects within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019]. The concept of heterogeneity is vitally important because it tells us how general our findings are likely to be and how much of the variance we see is the result of real biology or methodological differences (after meta-analytically accounting for sampling variance) [@Nakagawa2017; @Borenstein2019].

There are many phenomena driving effect heterogeneity. First, there is inherent uncertainty in estimating the ‘true’ population effect (e.g., a correlation coefficient) from smaller samples [@Borenstein2019]. A formal meta-analysis weights sample estimates by their precision, and removes this source of heterogeneity [@Gurevitch2018; @Koricheva2013]. Removing sampling variance is possible because we know, mathematically, how to calculate it for many effect measures. While unweighted analyses are common, weighting studies makes overall estimates more precise, and less susceptible to publication bias (see below), even if the overall mean itself is unbiased [@Morrissey2016a]. Second, there are sociological factors impacting heterogeneity. These include the ease with which novel and significant results are published relative to non-significant results, manifesting as publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, real biological processes can drive effect heterogeneity. This is particularly true in comparative physiology where we synthesise data from different species with varying life-histories, habitats, mating systems and reproductive modes. As comparative physiologists, this is the stuff that gets us up every morning! We can model these biological factors by including relevant predictors (moderator variables) and/or random effects (to account for non-independence of effects). By doing so, we can test predictions from hypotheses that we believe are key players impacting mean effect size and direction. This improves our understanding of the biological world around us and helps inform on future research directions.

Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it is important we know and understand this to determine how experiments can be designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' [following from @Cooperetal2019] -- factors we know vary, but are rather unsurprising to us (Figure \@ref(fig:fig1)). These might include differences in temperature, pH, dosages, water potential etc. Regardless of the relative importance of such variables compared to core questions in a meta-analysis, nuisance variables may complicate the interpretation of effect sizes used if not accounted for properly.

@@

Besides sampling variance, all the sources of heterogeneity discussed above contribute (to varying extents) to the 'apples and oranges' problem often touted as compromising the reliability of meta-analyses. Combining standardised effects, while ignoring these sources of heterogeneity can indeed weaken the reliability of a meta-analysis. However, many argue that heterogeneity is just what we are interested in synthesizing [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. In the words of Gene Glass, the founder of meta-analysis: "Of course it [meta-analysis] mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. Whether comparing apples and oranges is a good idea simply boils down to the specific question of interest, the population one wishes to make conclusions about and how heterogeneous effects actually are. If we are interested in generalising only to apples, then mixing fruit will not be ideal. However, if we want to understand why different 'fruit' vary in their properties, then we need to consider apples and oranges together [@Borenstein2019].

We tend to agree with @Glass2015, and his sentiment is particularly relevant to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms, including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are different. Synthesizing diverse study outcomes allows us to also understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction), and approaches (e.g., short-term vs long-term studies) that can explain the diversity of effect outcomes observed. Doing so provides a rich set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included in a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility (or ideally should be) are such a prominent feature of meta-analysis [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant and may even be misleading. It may, therefore be more useful to understand overall effects in each of these groups separately, and assess the extent to which they differ.

@@

Before overviewing strategies to deal with nuisance heterogeneity, it is worth first understanding how comparative physiologists are currently conducting meta-analyses and other quantitative syntheses. As such, we conducted a literature survey to better understand the different types of effect sizes being used in comparative physiology, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled. After searching for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species, we asked: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to nuisance heterogeneity? 3) If so, was such variation accounted for in the effect size or during analysis? and finally, 4) how many analyses used 'weighted' meta-analytic models?

@@

We included comparative studies / meta-analyses that were focused on physiological questions. Narrative reviews and comments were not considered because they did not collate quantitative data from the literature. Studies needed to have collected physiological trait data, such as hormone levels, thermal physiological parameters (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress measures and consequences (e.g., DNA damage, teleomere attrition), performance measurements (e.g., heart rates, sprint speed, swim speed, growth rates), body composition (e.g., fatty acid profiles) and immune traits (e.g., cell function, antibodies). Studies were also included if they measured the relationship between physiological traits and body size (e.g., metabolic scaling), reproductive allocation, or survival across taxa. Studies on behavior were only included if the synthesis was explicitly interested in the proximate physiological drivers of behavioral variation. We also included meta-analyses that quantified plasticity in physiological traits in response to environmental variation. While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature because this is the major data source in quantitative syntheses. We excluded quantitative syntheses interested in phenology, bio-mechanics, sexual size dimorphism (i.e., except if the study was interested in sexual differences in physiological traits), those focused on validating physiological methodology, or characterising population genetic structure, gene transcription, and quantitative genetics. There were also meta-analyses that focused on colour evolution which varied in their goals but were most often not focused on physiological questions. As such, these studies were also excluded. We restricted our focus to animal studies because most plant synthesis papers were focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output, and greenhouse gas emissions).

@@

As can be seen in Table \@ref(tab:tablea), none of the effect measures control for 'nuisance heterogeneity'. For example, the magnitude of *lnRR* or *SMD* might depend on differences in the temperatures, dosages, pH, or water potential applied to each treatment. The same applies to modelling the means. For example, standard metabolic rate (SMR) is temperature-dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest itself). While we could model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling (Figure \@ref(fig:fig0)). In Box 1 we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in Box 1 that can be easily generalized to alternative effect size measures and cases. Below, we show how we can formalise a traditional effect size, such as $lnRR$, to capture this ‘nuisance variation’ using a common, well understood effect measure of interest to comparative physiologists ($Q_{10}$).

```{r, fig0, fig.cap="**Effect sizes from old to new**. Traditional effect sizes (top left) are somehow detached from more biologically relevant measures commonly used in comparative physiology (top right). By combining traditional effects with physiological parameters, and correcting the resulting quantities for nuisance parameters, we can derive a new generation of biologically relevant, easily interpretable effect measures for use in meta-analyses."}
image <- image_read(here::here("figs", "Fig0_v1_ML.png"))
image
```

@@

Assume that we are interested in understanding the overall effect of, say, testosterone (T) on offspring traits [e.g., see @Podmola2018]. We might extract data from studies manipulating T in bird eggs. Of course, these studies are diverse, spanning many species and applying different dosages of T. If the studies are experimental in nature, then a standardised effect size, such as $lnRR$ or *SMD*, might be applicable. We then might fit an overall multilevel meta-analytic model as follows: 
$$
\begin{split}
Y_{i} = \beta_{0} + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i} \\
a_{k[i]}\sim N(0, \sigma^{2}_{phylogeny}\textbf{A}), \\
sp_{k[i]}\sim N(0, \sigma^{2}_{species}\textbf{I}), \\
s_{j[i]}\sim N(0, \sigma^{2}_{study}\textbf{I}), \\
e_{i}\sim N(0, \sigma^{2}_{residual}\textbf{I}), \\
m_{i} \sim N(0, v_{i}\textbf{I}) \\
\end{split}
$$

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect size, $v_{i}$ (which is known as the sampling variance for the effect). This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Importantly, if we have effect sizes that are derived from the *same* sample of organisms (e.g., because traits are measured on the same sets of individuals), then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying $\textbf{I}$ to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation, such as differences in the methods applied, species differences, or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our question: across all the species and studies manipulating egg T, how much of an overall effect does T have on offspring traits relative to control conditions?

Of course, the overall effect does not provide us with a way to interpret its magnitude in the context of how much T has been applied in the treatments used in the sample of studies. The most we can say about it is that the effect is small, moderate or large given the sample of effect sizes extracted from this set of studies. In the next section, we show how to 'correct' the overall effect size so that its interpretation is more meaningful.

@@

where $\sum \beta_{j}x_{j}$ is simply the sum of all effects for all moderators (fixed effects), $x_{j}$. All other notation is the same as described above. Importantly, the variable $x_{j}$ could be any moderator collected from studies, including dosage, temperature, salinity, or pH. Some moderators will be at the level of the study (e.g., was the study experimental or observational in nature) but others could be taken at the effect size level (i.e., was the effect derived using a temperature of 23°C or 35°C). By including these moderators in a meta-regression model, the interpretation of the overall meta-analytic mean, $\beta_{0}$, changes. For example, assume we collected the change in *in ovo* testosterone dosage relative to a control group (i.e., in ng T / g yolk) for our meta-analysis on egg T effects on bird offspring traits in birds. If the untransformed dosage, $x_{j}$, is included in the model then the overall meta-analytic mean estimate, $\beta_{0}$, would be the mean effect of T on reproduction when the dosage difference between the treatment and control group is 0 ng T / g yolk (i.e., it would be reflective of unmanipulated eggs). Of course, this interpretation does not make much sense.

We can, however, apply some simple transformations to make the overall mean estimate more intuitive while also improving the interpretation of non-linear parameters and interactions estimated from the model [@Schielzeth2010; @GelmanHill2007]. One simple transformation would be to centre the variable $x_{j}$ by subtracting each value from the mean to create a new input variable (i.e., $c_{j}=x_{j}-\bar{x}$, where the subscript *j* denotes each individual effect size dosage). The new variable, $c_{j}$, is now centered on the mean and can replace $x_{j}$ in the meta-regression model. When the model is refit with $c_{j}$, the estimated $\beta_{0}$ can now be more intuitively interpreted as the overall mean effect of CORT on reproduction at an average dosage difference between treatment and control groups. Importantly, by centering, we retain the original units (ng T / g yolk) of the variable, which can be very useful to ease the interpretation of the effect magnitude (e.g., when the dosage difference is 10 ng T / g yolk). Given it is very common to estimate and compare overall meta-analytic means (i.e., $B_{0}$) in meta-analyses, this centering can be particularly useful in making these means more interpretable. Mean centering can even be extended to include other 'nuisance' variables that might creep into the dataset; the same centering can be done with a second variable, and the overall mean adjusted and interpreted in the context of this new variable along with dosage. Of course, if one is not interested in the magnitude of the effect, we can simply model mean reproductive output as a function of dosage (often what are referred to as 'arm-based' meta-analytic models). Here, one can model the mean for each group (i.e., control and treatment) and account for the known sampling variance associated with each group mean. Dosage, or other continuous variables, like temperature, can then be modeled (accounting for sampling variance) to understand how mean reproductive output varies. It is important to recognise that all means need to be comparable and that these models are more complex to fit. While they can be equivalent to 'contrast-based' models, they require complex interactions to be estimated [@Nakagawa2015].

Centering moderator variables, like temperature and dosage, not only provides a way to make the overall mean effect more interpretable in the face of treatment heterogeneity, but it also allows more complex relationships to be fit without compromising interpretations, such as when one includes non-linear parameters (e.g., higher-order polynomials) and interactions with other variables. In other words, linear / main effects can still be interpreted normally in the presence of non-linear / interactions fit in the model [@Schielzeth2010; @GelmanHill2007]. We provide examples in the online Supplementary Material (with R code) to show readers how the models we describe above can be fit, and how meta-analytic means can be adjusted for treatment heterogeneity.

#### *Heterogeneity Analysis: How Much Nuisance Heterogeneity Actually Exists?*

@@

We would like to thank the Editorial board (Craig Franklin, Michaela Handel and Charlotte Rutledge) for inviting us to contribute to this special issue along with feedback on previous drafts. We would also like to thank Megan Head for comments on a previous draft. DWAN, SMD, SN, ML and REO were supported by ARC Discovery Projects (DP210101152 to DWAN, DE180100202 to SMD, and DP200100367 to SN, ML and DWAN). PP and SB were supported by a UNSW Scientia PhD Scholarship.
